---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r preprocessing, echo=T}

unzip("activity.zip",overwrite=F)
dataRaw<-read.csv("activity.csv", header = T)
dataRaw$date<-as.Date(dataRaw$date, format = "%Y-%m-%d")
```
## What is mean total number of steps taken per day?

```{r histogram, echo=T}
## Find the total steps for each day
daily.sum<-aggregate(dataRaw$steps,by=list(dataRaw$date),FUN=sum)
## Give good names to each column
colnames(daily.sum)<-c("date","totalSum")
## Let's plot the histogram
hist(daily.sum$totalSum, xlab="Number of steps", main ="Frequency of total number of  daily steps" )
mean.raw<-mean(daily.sum$totalSum, na.rm=TRUE)
print(paste("Mean number of steps per day : ",mean.raw))
median.raw<-median(daily.sum$totalSum,na.rm =TRUE)
print(paste("Median number of steps per day: ",median.raw))
```

## What is the average daily activity pattern?
```{r daily_activity_pattern_raw, echo=TRUE}
activity.pattern<-aggregate(dataRaw$steps,by=list(dataRaw$interval),FUN=mean, na.rm=TRUE)
colnames(activity.pattern)<-c("interval","mean")
plot(activity.pattern$interval,activity.pattern$mean, type = "l", xlab = "Interval",ylab = "Mean",main = "Time series of daily activity pattern across all days")
## Find the maximum interval of average daily activity
activity.pattern[which.max(activity.pattern$mean),]
```

## Imputing missing values

## Checking where the missing values are:

```{r Before any changes, echo=TRUE}
summary(dataRaw)
```

Therefore only setps has missing values.  

We are going to replace "NA" values with the mean of each interval
```{r Imputing missing values, echo=TRUE}
data<-merge(dataRaw,activity.pattern, by="interval")
for(i in 1:length(data$steps)){
  if (is.na(data[i,2])) {
    #print(data[i,2])
    data[i,2]<-data[i,4] 
  }
}
summary(data)
``` 
We are ready to produce the new histogram and check also for differences in the mean and median

## What is the average daily activity pattern?
```{r daily_activity_pattern_processed, echo=TRUE}
daily.sum.input<-aggregate(data$steps,by=list(data$date),FUN=sum, na.rm=TRUE)
## Give good names to each column
colnames(daily.sum.input)<-c("date","totalSum")
## Let's plot the histogram
hist(daily.sum.input$totalSum, xlab="Number of steps", main ="Frequency of total number of  daily steps" )
mean.input<-mean(daily.sum.input$totalSum, na.rm=TRUE)
print(paste("Mean number of steps per day : ",mean.input))
median.input<-median(daily.sum.input$totalSum,na.rm =TRUE)
print(paste("Median number of steps per day: ",median.input))
```
## Are there differences in activity patterns between weekdays and weekends?

Adding the new factor variable
```{r Adding factor variable, echo=TRUE}
data$day<-weekdays(data$date)
weekdayNames<-c("Monday","Tuesday","Wednesday","Thursday","Friday")
data$factor.day<-factor(data$day %in% weekdayNames,levels=c(FALSE,TRUE), labels=c("weekend","weekday"))
```
Let's subset by the new factor variable:
```{r subsetting by factor variable, echo=TRUE}
dataWeekdays<-data[ which(data$factor.day=="weekday"),]
dataWeekends<-data[ which(data$factor.day=="weekend"),]
```


## What is the average daily activity pattern?

Now let's find the daily mean by time periods.

```{r new_daily_activity_pattern_by_time_period, echo=TRUE}
activity.pattern.weekday<-aggregate(dataWeekdays$steps,by=list(dataWeekdays$interval),FUN=mean, na.rm=TRUE)
colnames(activity.pattern.weekday)<-c("interval","mean")
activity.pattern.weekend<-aggregate(dataWeekends$steps,by=list(dataWeekends$interval),FUN=mean, na.rm=TRUE)
colnames(activity.pattern.weekend)<-c("interval","mean")
```
 A plot the time series for weekdays and weekends:
```{r plot, echo=T,fig.width=7, fig.height=8} 
par(mfcol=c(2,1))
plot(activity.pattern.weekday$interval,activity.pattern.weekday$mean, type = "l",xlab="", ylab="",main = "Weekdays")
plot(activity.pattern.weekend$interval,activity.pattern.weekend$mean, type = "l",xlab="", ylab="",main = "Weekend")
mtext("Interval", side=1, line=4, cex = 1.2)
par(oma=c(0,6,0,0))
mtext("Mean number of steps", side=2, line=5, cex=1.5, outer = T)
```